---
title: "HW 3 Data Science"
author: "Victoria Mello (vsm2118)"
date: "October 14, 2023"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1


#### Read in the data

```{r}
data("instacart")

instacart = 
  instacart |> 
  as_tibble()
```

#### Answer questions about the data

This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart |> select(product_id) |> distinct() |> count()` products found in `r instacart |> select(user_id, order_id) |> distinct() |> count()` orders from `r instacart |> select(user_id) |> distinct() |> count()` distinct users.

Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

Next is a plot that shows the number of items ordered in each aisle. Here, aisles are ordered by ascending number of items.

```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

Our next table shows the three most popular items in aisles `baking ingredients`, `dog food care`, and `packaged vegetables fruits`, and includes the number of times each item is ordered in your table.

```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week. This table has been formatted in an untidy manner for human readers. Pink Lady Apples are generally purchased slightly earlier in the day than Coffee Ice Cream, with the exception of day 5.

```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```


## Problem 2

##### Data Cleaning
```{r}
data("brfss_smart2010")

# Data cleaning
brfss_data_cleaned <- brfss_smart2010 %>%
    janitor::clean_names() %>% 
    filter(topic == "Overall Health") %>%
    filter(response %in% c("Excellent", "Very Good", "Good", "Fair", "Poor")) %>%
    rename(state = locationabbr) %>% 
    mutate(response = factor(response, levels = c("Poor", "Fair", "Good", "Very Good", "Excellent"), ordered = TRUE))

variable.names(brfss_data_cleaned)
```


##### In 2002 and 2010, which states were observed at 7 or more locations?
```{r}

# States observed at 7 or more locations in 2002
states_2002 <- brfss_data_cleaned %>%
  filter(year == 2002) %>%
  group_by(state) %>%
  summarize(count = n()) %>%
  filter(count >= 7) %>%
  arrange(desc(count))

# States observed at 7 or more locations in 2010
states_2010 <- brfss_data_cleaned %>%
  filter(year == 2010) %>%
  group_by(state) %>%
  summarize(count = n()) %>%
  filter(count >= 7) %>%
  arrange(desc(count))

knitr::kable(states_2002, caption =  "States with 7 or more observations in 2002", format = "markdown")
knitr::kable(states_2010, caption =  "States with 7 or more observations in 2010", format = "markdown")

```

In 2002 a total of 36 states were observed at 7 or more locations. The states observed at the most locations in 2002 were PA with 40, followed by MA and NJ both having 32. By 2010, the total number of states observed at 7 or more locations increased to 45. The states observed at the most locations in 2010 were FL with 164, followed by NJ with 76, and TX with 64 locations.


##### Dataset limited to Excellent responses and a spaghetti plot:
```{r}

excellent_avg <- brfss_data_cleaned %>%
  filter(response == "Excellent") %>%
  group_by(state, year) %>%
  summarize(avg_data_value = mean(data_value)) %>%
  ggplot(aes(x = year, y = avg_data_value, group = state, color = state)) +
  geom_line() +
  labs(
    title = "Average Data Value Over Time (Excellent Responses) by State",
    x = "Year",
    y = "Average Data Value"
  ) +
  theme_minimal()

excellent_avg
```
The plot provides a useful tool for visualizing and comparing trends in the average data value of "Excellent" responses between states over time. It allows for a comparison between states over the years 2002 to 2010. The average data value is represented by the y-axis and the range of the graph on the y-axis is 10%-35%. Year is represented on the x-axis with a range of 2002-2010. However, due to there being 50 states, the plot is quite busy and it is challenging to distinguish between the lines of individual states. One observation that can be made from the plot is that most states appear to be following a similar fluctuating trend for the average "Excellent" responses to the "Overall Health" topic over time. However, without specific data points or a clearer view, it's difficult to make more detailed observations or conclusions.


##### Two-panel scatterplot showing the distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State in the years 2006 and 2010:
```{r}
# Filter the dataset to include only NY State and the years 2006 and 2010
ny_data <- brfss_data_cleaned %>%
  filter(state == "NY" & year %in% c(2006, 2010) & response %in% c("Poor", "Fair", "Good", "Very Good", "Excellent"))

custom_colors <- c("NY - Bronx County" = "black", "NY - Erie County" = "blue", "NY - Kings County" = "green", "NY - Monroe County" = "purple", "NY - New York County" = "orange",
                   "NY - Nassau County" = "cyan", "NY - Queens County" = "brown", "NY - Suffolk County" = "yellow", "NY - Westchester County" = "pink")

# Create the scatter plot for 2006 and 2010 and facet by year
plot_combined <- ggplot(ny_data, aes(x = response, y = data_value, color = locationdesc)) +
  geom_point(size = 3, alpha = 0.6, position = position_jitter(w = 0.2, h = 0)) +   
  scale_color_manual(values = custom_colors) +  # Use the custom color palette
  labs(title = "Distribution of Data Value by Location in New York State",
       x = "Response",
       y = "Data Value",
       color = "Location") +
  theme_minimal() +
  facet_grid(year ~ .)

plot_combined
```
As seen in the two panel scatterplot showing the distribution of data_value for responses (“Poor” to “Excellent”) among counties in NY State for the years 2006 and 2010, the "Good" response option to the question "How is your general health?" was the most common response, followed by "Excellent", "Fair", and "Poor" in both years. Data from 2010 has information on a greater number of NY counties than the 2006 dataset does, including information for these additional counties: Bronx, Erie, and Monroe counties. In 2006, the county with the highest distribution of "Excellent" health was New York County, whereas in 2010 it was Westchester county. In 2006, the county with the highest distribution of "Poor" health was also New York County, although in 2010 it was Kings county. 


## Problem 3 
```{r}
# Load demographic data
demographics <- read.csv("nhanes_covar.csv", skip = 4)
view(demographics)

# Load accelerometer data
accelerometer <- read.csv("nhanes_accel.csv") %>% 
  janitor::clean_names()
view(accelerometer)

# Tidy demographic data
demographics_clean <- demographics %>%
  janitor::clean_names() %>%
  drop_na() %>%
  filter(age >= 21) %>%
  mutate(education = case_when(
    education == 1 ~ "less than high school",
    education == 2 ~ "high school equivalent",
    education == 3 ~ "more than high school"
  ),
  sex = case_when(
    sex == 1 ~ "male",
    sex == 2 ~ "female"
  ))
view(demographics)

# Tidy accelerometer data and inner join
 joined_data = 
  accelerometer %>% 
  inner_join(demographics_clean, by = "seqn") %>% 
  relocate(seqn, sex, age, bmi, education) %>% 
  filter(age >= 21) %>% 
  mutate(
    sex = forcats::fct_relevel(sex, "male", "female"),
    education = forcats::fct_relevel(education, "less than high school", "high school equivalent", "more than high school"))

View(joined_data)
```


##### Table for the number of men and women in each education category, and a visualization of age distributions for men and women across education categories:
```{r}
# Create sex v education table
education_summary <- joined_data %>%
  group_by(education, sex) %>%
  summarize(
    count = n()
  ) %>%
  ungroup() 

knitr::kable(education_summary, caption =  "Number of Men and Women in each Education Category", format = "markdown")

# Bar chart depicting table 
education_summary %>%
  ggplot(aes(x = education, y = count, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Men and Women in Each Education Category", x = "Education Category", y = "Count") +
  theme_minimal()

# Create a histogram for age distributions by gender and education category
age_distribution_plot <- joined_data %>%
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  facet_grid(sex ~ education) +
  labs(title = "Age Distributions by Gender and Education Category", x = "Age", y = "Count") +
  theme_minimal()

# View the table and plots
education_summary
age_distribution_plot
```
The table illustrates the distribution of participants based on their education level and gender. This table offers insights into the number of participants in each education level, categorized by their gender. In the "Less Than High School Education" category, there are slightly more females than males, with 28 females compared to 27 males. For "High School Equivalent" there is a higher representation of males than females, with 35 males compared to 23 females. Finally, in the "More Than High School" category, there are slightly more females than males, with 59 females compared to 56 males. Looking at the plot of age distributions for men and women across education categories, it appears that the age distribution for both men and women in the "less than high school" category is wide, however, the distribution for women appears to be slightly more skewed towards older ages compared to men. In the "high school equivalent" category, the age distribution for both genders is broad, but the peak for women seems to be at a slightly older age than for men. Finally, in the "more than high school" category, the peak for women is at a younger age compared to men, indicating that women are more likely to complete higher education at a younger age than men.


##### Plot of total activity by age, sex, and education level:
```{r}
activity_total =
  joined_data %>% 
  rowwise() %>% 
  mutate(
    total_activity = sum(c_across(min1:min1440)) %>% 
    as.numeric())
        
# Plot total activity by age, gender, and education level
total_activity_plot <- ggplot(activity_total, aes(x = age, y = total_activity)) +
  geom_point(aes(color = sex)) +
  facet_wrap(~education) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Total Activity by Age, Gender, and Education Level",
       x = "Age",
       y = "Total Activity",
       color = "Gender") +
  theme_minimal()

total_activity_plot
```
Based on the scatter plot of total activity by age, sex, and education level, it appears that total activity for both genders tends to decrease with age across all education categories. Women tend to have higher total activity than men in the younger age groups, especially in the "less than high school" category. The "less than high school" category shows higher total activity for both genders compared to the other categories, especially for the younger age groups. Finally, the steepness of the graph trendlines suggests that people with less education tend to have less physical activity with increasing age. Specifically, the "less than high school" category shows a steeper decline in total activity with increasing age compared to the other categories, indicating that people with less education may be more susceptible to physical inactivity as they age


##### Three-panel plot that shows the 24-hour activity time courses for each education level and by sex:
```{r}
# Reshape the data into long format
hourly_data <- joined_data %>%
  pivot_longer(cols = starts_with("min"), names_to = "minute", values_to = "activity") %>%
  mutate(
    minute = as.numeric(gsub("min", "", minute)),  # Extract minute number
    hour = (minute - 1) %/% 60  # Calculate the hour
  ) %>%
  group_by(education, sex, hour) %>%
  summarize(mean_activity = mean(activity)) %>%
  ungroup()

# Create the three-panel plot
hourly_data %>%
  ggplot(aes(x = hour, y = mean_activity, color = sex)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, size = 1) +  
  facet_wrap(~education, scales = "free") +
  labs(x = "Hour of the Day", y = "Mean Activity Level") +
  theme_minimal()

hourly_data

```
Based on the results of the three-panel plot showing the 24-hour activity time courses for each education level by sex, it seems that overall, males have higher mean activity levels than females at every hour of the day. The only exception is for the "less than high school" group where the female mean activity level trend line is higher than that of males during the late afternoon and early evening hours. Additionally, individuals with more than a high school education have higher mean activity levels throughout the day compared to those with lower levels of education. Generally, it seems that mean activity levels are highest in the morning (around 8-10am) and gradually decrease throughout the day, with the lowest levels around 8-10pm.
