---
title: "HW 3 Data Science"
author: "Victoria Mello (vsm2118)"
date: "October 14, 2023"
output: github_document
---

# Promblem 1

```{r}

library(tidyverse)
library(p8105.datasets)
data("instacart")

```

Exploratory Analysis: 

The distribution of orders across the days of the week and the hours of the day...
```{r}
instacart %>%  
  group_by(order_dow, order_hour_of_day) %>% 
  summarize(n_orders = n()) %>% 
  ggplot(aes(x = order_dow, y = order_hour_of_day, fill = n_orders)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(x = "Day of the Week", y = "Hour of the Day", fill = "Number of Orders",
       title = "Distribution of Orders Across Days of the Week and Hours of the Day")
```

Top 10 most ordered products...
```{r}
instacart %>%
  group_by(product_name) %>%
  summarize(n_orders = n()) %>%
  arrange(desc(n_orders)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(product_name, n_orders), y = n_orders)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Product", y = "Number of Orders", title = "Top 10 Most Ordered Products")
```

The Instacart dataset has `r nrow(instacart)` observations and `r ncol(instacart)` variables. It provides detailed information about the orders placed by Instacart users, including the order sequence, the day and hour of the order, and the time since the previous order. The dataset also includes information about the products ordered, such as the product name and the aisle and department where the product is located.
